Domain-Driven, Generative Data Model for BigPetStore

### 2014-07-22 bdcloud_paper/latex/paper.tex
Edits of simulation text

\subsection{Relational Data Model}
The model generates data for four types of records: stores, customers, transactions, and transaction items (Figure~\ref{fig:relational-data-model}).  Store records contain both a unique identifier and a location in the form of a 5-digit zip code. Customer records consist of a unique identifier, a name, and a location given as a 5-digit zip code. Transaction records consist of a customer identifier, a store identifier, and a time. Transaction item records contain a transaction identifier and an item description.  The item description is a list of key-value pairs stored as a string.  Key-value pairs are used since item characteristics differ depending on the type of item.  For example, dog food has a brand, a flavor, a size (in lbs), and a per-unit cost, while poop bags have a brand, a color, a count, and a per-unit cost.

@@

\subsection{Generation of Stores}

Store records contain both a unique identifier and a location in the form of a 5-digit zip code.  The locations of the stores are modeled by a probability density function (PDF) that gives the probability that a store's location is the given zip code. We designed the PDF to give to zip codes in high-population, high-income areas the highest probabilities. The PDF is composed of two individual PDFs. One PDF determines the probability of each zip code as the population of that zip code over the total population:

\begin{equation}
p(\text{location}=z | \text{population}(z)) = \frac{\text{population}(z)}{\sum_{i} \text{population}(i)}
\end{equation}

@@

\begin{equation}
p(\text{location}=z | \text{income}(z)) = s ^ {\big( \frac{\text{income} - \min_i{\textrm{income}(i)}} {\max_i{\textrm{income}(i)} - \min_i{\textrm{income}(i)}} - 1 \big)}
\end{equation}

@@

\begin{align}
p(\text{location}=z) = &Z^-1 p(\text{location}=z | \text{population}(z)) \nonumber \\
&p(\text{location}=z | \text{income}(z))
\end{align}

where $Z$ is the normalization factor.  Given the small size ($\approx$30,000 zip codes) of the data set, $Z$ can be found directly by iterating over all of the zip codes and summing the scores.

The PDFs are parameterized using external data sets.  The zip codes' latitudes and longitudes came from the Zip Code Database Project \cite{Zips}.  The population and household income data for the zip codes were taken from from the U.S. Census American Community Survey \cite{ACS}.

The stores' locations are generated by sampling zip codes from the PDF using a Roulette Wheel sampler.

\subsection{Generation of Customers}
Customer records consist of a unique identifier, a name, a location given as a 5-digit zip code, and the number and species of their pets.

@@

For each proposed customer location zip code $z$, we compute the distance between $z$ and every store's zip code to find the closest store.  We compute the distance $d$ between $z$ and the zip code $s_z$ of the closest store.  The weight $w_z$ for each zip code $z$ is modeled using an exponential distribution with the average distance $\beta$.

\begin{align}
&w_z = \beta^{-1} \exp(-\beta^{-1} \, d_m(z)) \\
&d_m(z) = \min_{s_z} \, d(z, s_z) \nonumber
\end{align}

\begin{align}
&p(\text{location} = \text{z}) = W^{-1} w_z \\
&W = \sum_{z} w(z) \nonumber
\end{align}

@@

\begin{align}
N_p \sim U(1, b) \\
N_d \sim U(0, N_p) \nonumber \\
N_c = N_p - N_d \nonumber
\end{align}

\subsection{Simulation of Transactions}
A transaction model is created for each customer and integrated over the simulation's time frame to generate transactions. Transaction records consist of a customer, a store, a date and time, and a list of items purchased and their prices.  The store is set to the store located closest to the customer. The date and time is given as a floating-point number with units of days since the start of the simulation.  (Time zones are not taken into account.)  Transaction times are generated from a Monte Carlo process based on the customer will use up items such as dog food (Section~\ref{sec:transaction-times}).  The purchased items are generated by a Hidden Markov Model (HMM) parameterized by the transaction time and amount of time remaining for the items in the customer's inventory (Section~\ref{sec:transaction-purchases}).

@@

\subsubsection{Simulation of Transaction Times} \label{sec:transaction-times}

Transaction times are simulated using a Monte Carlo method (Figure~\ref{fig:trans_sim}).  The usage over time of each category is simulated, from which transaction times are found for each item category. The earliest transaction time is taken as the proposed transaction time.  The probability of the transaction time is calculated using a PDF.  If rejected, new transaction times are proposed.  Otherwise, the purchased items are modeled using a separate process (discussed below). After the items are chosen, the process begins again with the simulation of the item category usages.

@@

\begin{equation}
\frac{da_i}{dt} = -\mu_i + \sigma_i\, dW_i(t)
\end{equation}

where $a_i$ is the amount of item category $i$ ``stuff'' remaining at time $t$. The variable $u_i$ gives the average usage rate and $\sigma^2_i$ gives the variance of the usage rate for item category $i$. The SDE is numerically integrated using the Euler-Maruyama method\cite{Klouden13} with time step $\Delta t_n$ sampled from a Poisson process:

\begin{align}
&a_{n+1, i} = a_{n,i} - \mu_{i, c} \, \Delta t_n + \sigma_{i, c} \, \sqrt{\Delta t_n} R_n \\
&\Delta t_n \sim \text{Exp}(\lambda_i) \nonumber \\
&R_n \sim \text{N}(0, 1) \nonumber
\end{align}

where $\lambda_i$ is the item category's average usage rate, $\mu_i$ is the item category's average amount used per use, and $\sigma^2_i$ is the variance of the amount of the item category used per use. The parameter $\lambda_i$ is defined for each item category.  The parameters $\mu_{i, c}$, and $\sigma^2_{i,c}$ are found based on a base rate for each item category multiplied by the number of pets of the appropriate species the customer has.  The exhaustion time $T_{E,i}$ for each item category is found by simulating the usage until $a_n = 0$.

The proposed transaction time is found from the exhaustion times.  For each item category, the offset time $T_{O, i}$ between when a customer would visit the store and the exhaustion time is sampled from an exponential distribution parameterized separately for each customer $c$. A proposed transaction time $T_{P, i}$ for each item category is found by subtracting the offset time $T_{O, i}$ from the exhaustion time $T_{E, i}$.  The earliest proposed transaction time is taken as the overall proposed transaction time $T_T$.

\begin{align}
&\beta_c \sim \text{U}(a, b) \\
&T_{O, i} \sim \text{Exp}(\beta^{-1}_c) \\
&T_{P, i} = T_{E,i} - T_{O, i} \\
&T_T = \min_i \, \{  T_{P, i}\}
\end{align}

The acceptance probability $p(t=T_T)$ of the proposed transaction time $T_T$ is modeled using a PDF that incorporates factors such as store hours.  The PDF could be extended to incorporate additional factors including time-dependent events such as sales, weather, and the amount of money the customer has available at the time.

\begin{equation}
p(t=T_T) = p(t=T_T | \text{store hours})
\end{equation}

The proposed transaction time is accepted if $p(t=T_T) < r$, where $r \sim \text{U}(0, 1)$. Until the proposed transaction time is accepted, a new proposed transaction time is generated by sampling new offset times. If the proposed transaction time is accepted, the shopping cart of items is generated through a separate simulation discussed in Section~\ref{sec:transaction-purchases} below. Transactions are generated until the accepted transaction time is later than the end time given as a simulation parameter.

@@

\subsubsection{Shopping Cart Items} \label{sec:transaction-purchases}

A customer's purchases in each transaction are modeled using a layered Hidden Markov Model (HMM) (Figure~\ref{fig:shopping_cart_sim}).  The HMM has three states: the start state, the purchase states, and the stop state. There are an infinite number of purchase states, identified by the exhaustion times of the customer's item categories, the transaction time, and the number of items purchased in the current transaction. The observables of the states correspond to the item categories.  Each item category $i$ is assigned a weight $w_i$ according to the amount of time between the transaction time $T_t$ and $T_{E,i}$ when each item category will be exhausted (based on the usage simulations):

\begin{equation}
w_i = - \lambda_i \exp(-\lambda_i (T_{E, i}- T_{T}))
\end{equation}

@@

\begin{align}
&p(X=i) = W^{-1} w_i \\
&W = \sum_i w_i \nonumber
\end{align}

@@

\begin{align}
Pr(X_{n+1}&=x|X_n=y) = \\
& \left\{ 
  \begin{array}{l l}
   (1.0 - w_l) W^{-1} \sum_f w_f w_{f,s}(x, y)  & \quad \text{if $x \neq y$}\\
   w_l & \quad \text{if $x = y$}
  \end{array} \right.
\end{align}

\begin{equation}
w_{f,s}(x,y) = \left\{ 
  \begin{array}{l l}
   w_{f,s}  & \quad \text{if $x$.$f$ = $y$.$f$}\\
   1 - w_{f, s} & \quad \text{if $x$.$f$ $\neq$ $y$.$f$}
  \end{array} \right.
\end{equation}

\begin{align}
&w_l \sim N(\mu, \sigma^2) \\
&w_f \sim N(\mu, \sigma^2) \\
&w_{f, s} \sim N(\mu, \sigma^2)
\end{align}

\begin{equation}
W = \sum_i \sum_{j \neq i} \sum_f w_f w_{f,s}(i, j)
\end{equation}

>>>>

\subsection{Relational Data Model}
The model generates data for four types of records: stores, customers, transactions, and transaction items (Figure~\ref{fig:relational-data-model}).  Store records contain both a unique identifier and a location in the form of a 5-digit zip code. Customer records consist of a unique identifier, a name, and a location given as a 5-digit zip code. Transaction records consist of a customer identifier, a store identifier, and a time given in days since the beginning of the simulation. Transaction item records contain a transaction identifier and an item description.  The item description is a list of key-value pairs stored as a string.  Key-value pairs are used since item characteristics differ depending on the type of item.  For example, dog food has a brand, a flavor, a size (in lbs), and a per-unit cost, while poop bags have a brand, a color, a count, and a per-unit cost.

@@

\subsection{Generation of Stores}
The locations of the stores are modeled by a probability density function (PDF) that gives the probability that a store's location is the given zip code. We designed the PDF to give to zip codes in high-population, high-income areas the highest probabilities. The PDF is composed of two individual PDFs. One PDF determines the probability of each zip code as the population of that zip code over the total population:

\begin{equation*}
p(\text{location}=z | \text{population}(z)) = \frac{\text{population}(z)}{\sum_{i} \text{population}(i)}
\end{equation*}

@@

\begin{equation*}
p(\text{location}=z | \text{income}(z)) = s ^ {\big( \frac{\text{income} - \min_i{\textrm{income}(i)}} {\max_i{\textrm{income}(i)} - \min_i{\textrm{income}(i)}} - 1 \big)}
\end{equation*}

@@

\begin{align*}
p(\text{location}=z) = &Z^-1 p(\text{location}=z | \text{population}(z)) \\
&p(\text{location}=z | \text{income}(z))
\end{align*}

where $Z$ is the normalization factor.  Given the small size ($\approx$30,000 zip codes) of the data set, $Z$ can be found directly by iterating over all of the zip codes and summing the scores. The population and household income data for the zip codes were taken from from the U.S. Census American Community Survey \cite{ACS}.  The stores' locations are generated by sampling zip codes from the PDF using a Roulette Wheel sampler.

\subsection{Generation of Customers}
For each proposed customer location zip code $z$, we compute the distance between $z$ and every store's zip code to find the closest store.  We compute the distance $d$ between $z$ and the zip code $s_z$ of the closest store.  The zip codes' latitudes and longitudes came from the Zip Code Database Project \cite{Zips}.  The weight $w_z$ for each zip code $z$ is modeled using an exponential distribution with the average distance $\beta$.

\begin{align*}
&w_z = \beta^{-1} \exp(-\beta^{-1} \, d_m(z)) \\
&d_m(z) = \min_{s_z} \, d(z, s_z)
\end{align*}

\begin{align*}
&p(\text{location} = \text{z}) = W^{-1} w_z \\
&W = \sum_{z} w(z) \nonumber
\end{align*}

@@

@@

\begin{align*}
N_p \sim U(1, b) \\
N_d \sim U(0, N_p) \\
N_c = N_p - N_d
\end{align*}

\subsection{Simulation of Transactions}
A transaction simulation is run for each customer. The transaction's store is set to the store located closest to the customer.  Transaction times are generated from a Monte Carlo process that proposes transaction times based on the usage of the customer's items and Poisson processes modeling the amount of time between the customer visiting the store and running out of the items (Section~\ref{sec:transaction-times}).  The purchased items are generated by a Hidden Markov Model (HMM) parameterized by the transaction time and amount of time remaining for the items in the customer's inventory (Section~\ref{sec:transaction-purchases}).

@@

\subsubsection{Simulation of Transaction Times} \label{sec:transaction-times}
Transaction times are simulated using a Monte Carlo method (Figure~\ref{fig:trans_sim}).  The usage over time of each item category is simulated. The time between the customer's visit to the store to buy more items in each category and the exhaustion time of that item category is modeled using a Poisson process. Proposed transaction times for each item category are computed based on the exhaustion time and time sampled from the Poisson process. The earliest transaction time is taken as the overall proposed transaction time.  The probability of the transaction time is calculated using a PDF.  If rejected, new transaction times are proposed.  Otherwise, the purchased items are modeled using a separate process (discussed below). After the items are chosen, the process begins again with the simulation of the item category usages.

@@

\begin{equation*}
\frac{da_i}{dt} = -\mu_i + \sigma_i\, dW_i(t)
\end{equation*}

where $a_i$ is the amount of item category $i$ ``stuff'' remaining at time $t$. The variable $u_i$ gives the average usage rate and $\sigma^2_i$ gives the variance of the usage rate for item category $i$. The SDE is numerically integrated using the Euler-Maruyama method\cite{Klouden13} with time step $\Delta t_n$ sampled from an exponential distribution:

\begin{align*}
&a_{n+1, i} = a_{n,i} - \mu_{i, c} \, \Delta t_n + \sigma_{i, c} \, \sqrt{\Delta t_n} R_n \\
&\Delta t_n \sim \text{Exp}(\lambda_i) \\
&R_n \sim \text{N}(0, 1)
\end{align*}

where $\lambda_i$ is the item category's average usage rate, $\mu_i$ is the item category's average amount used per time used, and $\sigma^2_i$ is the variance of the amount of the item category used per time used. The parameter $\lambda_i$ is defined for each item category.  The parameters $\mu_{i, c}$, and $\sigma^2_{i,c}$ are found based on a base rate for each item category multiplied by the number of pets of the appropriate species the customer has.  The exhaustion time $T_{E,i}$ for each item category is found by simulating the usage until $a_n = 0$.

The proposed transaction time is found from the exhaustion times (Eq.~\ref{eq:transaction-times}).  For each item category, the offset time $T_{O, i}$ between when a customer would visit the store and the exhaustion time is sampled from an exponential distribution parameterized separately for each customer $c$. A proposed transaction time $T_{P, i}$ for each item category is found by subtracting the offset time $T_{O, i}$ from the exhaustion time $T_{E, i}$.  The earliest proposed transaction time is taken as the overall proposed transaction time $T_T$.

\begin{align} \label{eq:transaction-times}
&\beta_c \sim \text{U}(a, b) \nonumber \\
&T_{O, i} \sim \text{Exp}(\beta^{-1}_c) \nonumber \\
&T_{P, i} = T_{E,i} - T_{O, i} \nonumber\\
&T_T = \min_i \, \{  T_{P, i}\}
\end{align}

The acceptance probability $p(t_{n+1}=T_T|t_n)$ of the proposed transaction time $T_T$ is modeled using a PDF (Eq.~\ref{eq:transaction_time_pdf}). For now, the PDF only ensures that the proposed transaction time is more recent than the last transaction time.

%The PDF could be extended to incorporate additional factors including time-dependent events such as sales, weather, and the amount of money the customer has available at the time.

\begin{equation} \label{eq:transaction_time_pdf}
p(t_{n+1}=T_T|t_n) = \left\{ 
  \begin{array}{l l}
   1 & \quad t_{n+1} \geq t_n\\
   0 & \quad t_{n+1} < t_n
  \end{array} \right.
\end{equation}

The proposed transaction time is accepted if $p(t_{n+1}=T_T) < r$, where $r \sim \text{U}(0, 1)$. Until the proposed transaction time is accepted, a new proposed transaction time is generated by sampling new offset times. If the proposed transaction time is accepted, the shopping cart of items is generated through a separate simulation discussed in Section~\ref{sec:transaction-purchases} below. Transactions are generated until the accepted transaction time is later than the end time given as a simulation parameter.

@@

\subsubsection{Simulation of Transaction Items} \label{sec:transaction-purchases}

A customer's purchases in each transaction are modeled using a layered Hidden Markov Model (HMM) (Figure~\ref{fig:shopping_cart_sim}).  The HMM has three types of states: the start state, the purchase states, and the stop state. There are an infinite number of purchase states, identified by the exhaustion times of the customer's item categories, the transaction time, and the number of items purchased in the current transaction. The observables of the states correspond to the item categories.  Each item category $i$ is assigned a weight $w_i$ according to the amount of time between the transaction time $T_t$ and $T_{E,i}$ when each item category will be exhausted (based on the usage simulations):

\begin{equation*}
w_i = - \lambda_i \exp(-\lambda_i (T_{E, i}- T_{T}))
\end{equation*}

@@

\begin{align*}
&p(X=i) = W^{-1} w_i \\
&W = \sum_i w_i \nonumber
\end{align*}

@@

\begin{align*}
Pr(X_{n+1}&=x|X_n=y) = \\
& \left\{ 
  \begin{array}{l l}
   (1.0 - w_l) W^{-1} \sum_f w_f w_{f,s}(x, y)  & \quad \text{if $x \neq y$}\\
   w_l & \quad \text{if $x = y$}
  \end{array} \right.
\end{align*}

\begin{equation*}
w_{f,s}(x,y) = \left\{ 
  \begin{array}{l l}
   w_{f,s}  & \quad \text{if $x$.$f$ = $y$.$f$}\\
   1 - w_{f, s} & \quad \text{if $x$.$f$ $\neq$ $y$.$f$}
  \end{array} \right.
\end{equation*}

\begin{align*}
&w_l \sim N(\mu, \sigma^2) \\
&w_f \sim N(\mu, \sigma^2) \\
&w_{f, s} \sim N(\mu, \sigma^2)
\end{align*}

\begin{equation*}
W = \sum_i \sum_{j \neq i} \sum_f w_f w_{f,s}(i, j)
\end{equation*}
